---
title: "MetaboPeak"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MetaboPeak}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates a complete metabolomics workflow using the example datasets included in the package.
The data originate from a metabolomics experiment investigating the effects of elevated CO2, and drought on two tree species.
This is an informal and code heavy version of the upcoming MetaboPeak paper.

# LC-MS data and MetaboPeak

Liquid chromatography-mass spectrometry has emerged as a powerful analytical technique and it combines the separation capabilities of liquid chromatography with the detection and identification capabilities of mass spectrometry.
LC-MS workflow involves several steps, starting with sample preparation, followed by chromatographic separation, ionization, mass analysis, and data processing.
In this paper, we focus on data processing and the challenges associated with it.

The main role of MetaboPeak is to simplify and automate processes commonly done in Excel.
At the same time, it offers options to link files and filter data.
The whole package could be divided into 3 main groups of functions.
The first and largest are features aimed at better understanding the data, peak information useful in combination with chromatograms and linking peak data to treatment information.
The second group focuses on data filtering and the third on annotation and linking data to in-house library.

# Defining data for MetaboPeak

MetaboPeak is designed to work with processed data from a mass spectrometry software (MS-DIAL, MZmine, GeneData etc).
Every software has a little bit different output, however with few minor changes and saving it as csv the data should be fine to use with this package.

## Package data

- neg ... mass spectrometry processed data from MS-DIAL in negative mode
- pos ... mass spectrometry processed data from MS-DIAL in positive mode
- t_info ... table with treatment info
- pkg_lib ... table with in-house library with m/z and RT corresponding with LC-MS analyses

### Data structure

Most of the functions require mass spectrometry data as shown in example.
First column is the peak identification code, than there are the samples, mass to charge ration and retention time.
In rows are the peaks with abundances for the samples and their measured m/z and RT.

```{r, message=FALSE}
library(MetaboPeak) # if not working then go to working directory where the package is and then by devtools load_all(".")
head(neg)
```

Then there is the treatment information and in-house library, where for the treatment information there needs to be first column with the sample name, that correspond with the column names in previous data set and some treatment information in the next columns.
For the library data set, there should be m.z and RT columns and compound names.

```{r}
head(t_info)
head(pkg_lib)
```

For more information about each data set use help:

```{r}
?neg
```

# Preprocessing with MetaboPeak

The example dataset included with the package has already been processed using `treatGroup()`, which makes and unique code for combination of treatments.
When analyzing your own experiments, apply `treatGroup()` to your treatment table before using downstream functions.

## Isolating Biologically Relevant Windows Across Experimental Groups

In mass-spectrometry studies, raw chromatograms always contain baseline noise, solvent fronts at the very beginning of the run, and harsh matrix wash-off spikes at the absolute end. 
Trimming away these uninformative regions before performing complex statistical profiling is a critical step for a clean classification workflow.
The package handles this evaluation with the `interestRegion()` utility function.

Rather than forcing complex manual row indexing or data frame slicing syntax, this function scans your dataset and strips away early-eluting dead volumes and late-eluting background bleed. 
It drops columns or rows outside your boundaries and returns a clean, reduced data frame identifying only those specific peak features that fall within your stable, biologically active elution timeframe.

```{r}
# Reduce the original data frame to retention times that are biologically relevant
cleaned_data <- interestRegion(
  x = neg, 
  start = 1.4, 
  end = 25.0
)

# Print a summary statement comparing the original and cleaned dataset sizes
original_count <- nrow(neg)
cleaned_count  <- nrow(cleaned_data)
removed_count  <- original_count - cleaned_count

cat(paste0(
  "Data trimming complete.\n",
  "- Original dataset features: ", original_count, "\n",
  "- Retained biological features: ", cleaned_count, "\n",
  "- Background noise features removed: ", removed_count, ".\n"
))
```

## Tracking Feature Completeness Across Experimental Groups

In mass-spectrometry studies, determining whether a biomarker candidate is consistently expressed across your entire biological group cohort—or if it is compromised by zero markers and missing readings—is critical for rigorous classification.
The package handles this evaluation with the `treatCases()` utility function.

Rather than forcing complex manual row extraction loops, this function scans your abundance data within flexible numerical envelopes (`mz_tolerance` and `rt_tolerance`).
It maps individual samples directly to their treatment groups, scans for structural missing values, and returns a clean character vector identifying only those specific treatment categories that achieve a 100% complete case record for that peak.

```{r}
# Discover which experimental groups have 100% complete data for a specific feature
complete_cohorts <- treatCases(
  x = neg, 
  y = t_info, 
  mass = 447.222, 
  RT = 13.226, 
  n = 48,
  mz_tolerance = 0.01,
  rt_tolerance = 0.05
)

# Convert the result vector into a single, clean sentence for reporting
if (length(complete_cohorts) > 0) {
  groups_sentence <- paste(complete_cohorts, collapse = ", ")
  cat(paste("This feature is completely present in the following groups:", groups_sentence, ".\n"))
} else {
  cat("This feature does not have a 100% complete case profile in any group.\n")
}

```

# Peak information

Before filtering or performing downstream statistical evaluations, it is critical to understand the underlying distributions, signal variability, and missingness patterns within your metabolomics dataset.
The package provides a quick diagnostics tool named `peakInfo()` to generate structural profiles for your data features.

## Summarizing All Peaks

The function isolates your abundance values from trailing structural variables (like Mass-to-Charge Ratio and Retention Time) using the sample total count parameter (`n`).
It treats legacy zero markers as missing observation values (`NA`), computes fast matrix statistics, and outputs a complete, mathematically structured data frame.

```{r}
# Compute dataset profile statistics across 48 samples
dataset_profile <- peakInfo(neg, n = 48)

# View summary layout columns
head(dataset_profile, 5)
```

The returned dataset appends descriptive data tracks for each peak, including: \* **pctNAs**: The ratio of structural missing values relative to total sampling depth.
\* **Max**: Absolute peak ceiling intensity across standard groups.
\* **Mean & SD**: Rounding-controlled statistical center and variation points.
\* **CV**: The relative standard deviation metric ($SD / Mean$) indicating feature reliability.

## Inspecting an Individual Feature

If you notice an anomaly or want to trace the metrics of a specific biomarker candidate, you can extract its subset profile directly by defining the unique row label in the `ID` parameter:

```{r}
# Isolate profiling rows for a target peak identifier
peakInfo(neg, n = 48, ID = "254")
```

# Data filtering

Metabolomic datasets generated by mass spectrometry are typically massive, often containing thousands of peaks.
Many of these peaks represent background noise, unstable measurements, or redundant signals (such as isotopic patterns or adducts) that can decrease statistical power.
Filtering your data reduces this noise, shrinks file sizes, and ensures that down-stream statistical analyses are focused only on high-quality, biologically meaningful features.The package provides a comprehensive, multi-stage filtering function called `peakFilter()`.
It allows you to apply three distinct filtering strategies sequentially or independently.

## NA (missing values) filter

Missing values (recorded as NA or 0) are highly common in metabolomics due to compound concentrations falling below the instrument's limit of detection.
However, excessive missingness across a dataset makes a peak statistically unusable.

- How it works: This filter evaluates the percentage of missing values (NA) for every single peak (row) across the entire dataset.
  If a peak's total percentage of missing values exceeds your specified threshold, it is flagged for removal.

- The "Complete Case" Exception: To prevent the accidental deletion of important biomarkers, the filter includes a safety check.
  If a peak is completely present (has zero missing values) within even a single treatment group, it is saved and kept in the dataset.
  This ensures you do not throw away a compound that is completely absent in your control group but highly abundant in your treated group.

## CV (Coefficient of Variation) filter

The Coefficient of Variation (CV) measures technical variation and reproducibility.
It is calculated by dividing the standard deviation of a peak's intensity by its mean value ($\text{CV} = \sigma / \mu$).
A high CV means the peak's measurements are highly unstable and noisy.

- How it works: The function calculates the CV for each peak across all samples.
  If a peak's CV is higher than your defined threshold (e.g., 0.8 or 80% variation), it is flagged as unreliable.

- The "Complete Case" Exception: Just like the NA filter, the CV filter protects potentially critical group-specific features.
  If a peak has a perfectly complete set of non-missing values within at least one treatment group, it bypasses the CV cutoff and is safely retained.

## RT (Retention Time & Correlation) filter

In liquid chromatography-mass spectrometry (LC-MS), a single metabolite can generate multiple distinct peaks due to in-source fragmentation, isotopes, or different adducts.
These redundant peaks share the same retention time (RT) and their intensities fluctuate identically across samples.

- How it works: This filter groups peaks using a two-step network analysis approach:

  - Sliding RT Window: It rounds and groups peaks that fall within a tight Retention Time window (e.g., 0.02 minutes).
  - Spearman Correlation Graph: Within each RT window, it calculates the pairwise Spearman correlation of peak intensities across all samples. Pairs that exceed your correlation threshold (e.g., 0.95) are linked together to build an interconnected network graph.

- Peak De-duplication: For each interconnected cluster of redundant peaks, the algorithm keeps only one representative feature—the peak with the highest mass-to-charge ratio (m.z).
  All other highly correlated companion peaks within that cluster are safely removed, effectively de-duplicating your data.

### Quick Example

```{r}
# Apply all three filters: NA threshold 50%, CV threshold 80%, 
# and group peaks within 0.02 min RT windows with a 95% correlation cutoff.
filtered_neg <- peakFilter(
  x = neg, 
  y = t_info, 
  fNA = c('T', 0.5), 
  fCV = c('T', 0.8), 
  fRT = c('T', 3, 0.02, 0.90),
  parallel = TRUE
)

dim(neg)
dim(filtered_neg)
```

# Peak correlations

## Investigating Feature Relationships with Peak Correlation

During mass-spectrometry preprocessing, tracking specific feature relationships—such as looking for matching adducts, in-source fragments, or verifying cross-mode alignments between positive and negative ionization datasets—is a common workflow step.
The package provides a quick exploration utility called `checkCorr()` to evaluate whether two features share a synchronized chromatographic elution profile.

The function maps targeted Mass-to-Charge Ratio (`m.z`) targets within an explicit mass tolerance window (defaulting to `0.05` Da).
If multiple isobaric peaks match the profile window, the algorithm isolates the single feature tracking closest to your target Retention Time (`RT`), computes a profile correlation across all sample layers, and handles missing observations smoothly using complete pairwise steps.

### Example: Verifying Cross-Mode Co-Elution

```{r}
# Evaluate correlation between 2 peaks in the same dataset
correlation_result <- checkCorr(
  x = pos, 
  masses = c(595.2, 449.08), 
  RT = 11.8, 
  n = 48
)

print(paste("Feature Profile Correlation:", round(correlation_result, 3)))


# Evaluate correlation between a positive mode peak and a corresponding negative mode target
correlation_score <- checkCorr(
  x = pos, 
  y = neg, 
  masses = c(595.2, 593.15), 
  RT = 11.8, 
  n = 48,
  tolerance = 0.05
)

print(paste("Feature Profile Correlation:", round(correlation_score, 3)))

```

## Automated Co-Elution Screening Across Datasets

Manually querying individual peak relationships can become an exhausting task over massive metabolomics datasets.
To simplify discovery workflows, the package introduces an automated tracking screening engine called `findCorrelatedPairs()`.
This function parses your peak tables, builds structural sliding retention time matrices, and isolates features that share both a tight Retention Time proximity (`rt_window`) and a highly synchronized intensity pattern across your samples.

This function can run internally on a single dataset to uncover candidate isotope clusters or in-source fragments, or scan cross-mode datasets (comparing positive and negative screens) to find matched molecular ion partners.

```{r}
# Scan for related features co-eluting within 0.04 minutes sharing a correlation >= 0.90
highly_linked_features <- findCorrelatedPairs(
  x = pos, 
  y = neg, 
  n = 48, 
  rt_window = 0.04, 
  cor_threshold = 0.90
)

# Inspect the top strongly correlated feature links
head(highly_linked_features)
```

The output yields an ordered, high-utility matrix documenting the identity codes, respective mass coordinates, specific retention run times, exact absolute chromatographic time splits (`RT_Diff`), and calculated Spearman scores sorted from strongest relationship downwards.

## Visualizing Feature Co-Elution Profiles

Once you have identified interesting candidate pairs using `findCorrelatedPairs()`, it is highly recommended to visually inspect their abundance tracking patterns across your sample cohort.
This helps rule out false positives driven by localized sample artifacts.

The package provides a 1-line plotting tool called `plotPeakCorrelation()`.
It extracts the target feature arrays, normalizes their intensity profiles (allowing you to cleanly compare a high-abundance parent ion with a low-abundance isotope or fragment side-by-side), and draws their trends across all samples.

```{r}
# Plot the alignment profile of a discovered high-correlation peak pair
plotPeakCorrelation(
  x = pos, 
  y = neg, 
  peak_id_x = "941", 
  peak_id_y = "2063", 
  n = 48
)
```

This generates a publication-ready line graphic overlaying both peak trends, automatically calculating and engraving the active Spearman correlation score, mass coordinates, and elution times straight onto the figure canvas header.

# Background noise variability check

## Visualizing Feature Stability via Variance Barplots

When picking biomarker candidates or doing quality control checks on target compound groups, it is incredibly informative to see how feature abundances are distributed across your sample cohort.
Features with ultra-low variance are typically highly dependable signals.
The package offers an express plotting function called `barPeaks()` to automatically discover and plot the most stable signals tracking near a target mass window.

The function screens all records matching your query within a custom Daltons window (`tolerance`), calculates their Coefficients of Variation (CV), and selects up to 3 features exhibiting the lowest relative variance.
It then stacks their absolute abundance trends into a clean multi-panel graphic for quick evaluation.

```{r}
# Find and plot profiles for the 3 most stable peaks around m/z 355.10
stable_features <- barPeaks(
  x = neg, 
  mass = 355.10, 
  n = 48, 
  tolerance = 0.02
)

# Inspect returning data summaries
print(stable_features$peak_names)
```

The function returns a useful descriptive payload containing the extracted raw sample arrays, individual CV metrics, and the precise bar mid-point configurations, while dynamically safeguarding your default plotting console dimensions.

## Tracking Maximal Feature Abundance

When analyzing profiling runs, it is often critical to find which specific biological sample or treatment condition expresses a compound at its absolute highest intensity.
This is particularly useful for identifying sample outliers, discovering positive control behaviors, or tracking down which sample to use for downstream targeted fragmentation (MS/MS) experiments.
The package provides a direct query tool called `highestAbund()`.

Instead of forcing you to subset your arrays manually, the function scans the peak matrix within a reliable mass tolerance window (`tolerance`).
If multiple chromatographic features share that mass space, it searches across all of them simultaneously to isolate the single highest intensity coordinates and returns the exact text name of the sample it belongs to.

```{r}
# Find which sample column contains the highest intensity for mass 593.128
top_expressing_sample <- highestAbund(
  x = neg, 
  mass = 593.128, 
  n = 48,
  tolerance = 0.01
)

print(paste("Maximum intensity observed in:", top_expressing_sample))
```

# Working with library

## Annotating Experimental Features Against Standard Reference Libraries

Once structural peak reduction filters are complete, identifying candidate compound annotations by referencing data against an in-house database or a target chemical library is standard practice.
The package achieves this using the `libraryHits()` database exploration engine.

The function maps your experimental mass peaks (`m.z`) and chromatography timelines (`RT`) directly against standard standard entries within customizable accuracy parameters (`mz.range` and `RT.range`).
If structural duplication issues surface (where one reference target captures multiple near experimental peaks), an automated proximity step (`unique = TRUE`) filters down to keep only the absolute closest match.
All other unassigned peaks are marked as `"unknown"`.

```{r}
# Match raw screen records against a validated database data frame
annotation_payload <- libraryHits(
  x = neg, 
  l = pkg_lib, 
  RT.range = 0.20, 
  mz.range = 0.005, 
  unique = TRUE
)

# Extract total count of unique identities matched
annotated_count <- length(annotation_payload$unique_names)

# Convert the tracking result into a clean summary sentence
if (annotated_count > 0) {
  cat(paste("Library matching complete: successfully assigned identity tags to", annotated_count, "metabolite compounds.\n"))
} else {
  cat("Library search finished: no peaks matched the standard library criteria.\n")
}
```

### Structure of the Returned List Object

The `libraryHits()` function returns a named list with a length of 3.
Below is the technical data schema of the output object:

- **`List Element 1: lib_small`** (Data Frame) Clean annotation report showing only successful assignments and their retention time drift (`RTdiff`).

- **`List Element 2: lib_all`** (Data Frame) Matrix mirroring your entire input dataset, perfect for passing directly into downstream statistical tests.

- **`List Element 3: unique_names`** (Character Vector) A simple list of found metabolites, ideal for pathway enrichment analysis or clean text reporting.
